---
title: "Nova Scotia Employee Absenteeism - in Python"
author: "Chris Selig"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = FALSE,
    warning = FALSE,
    message = FALSE
    )
```

```{r}
library(reticulate)
```


```{r setEnvironment}
use_condaenv("learnPython", required = TRUE)
```

```{r}
# conda_install('learnPython','pretty_html_table')
```

```{python}
import os
import pandas as pd
import matplotlib.pyplot as plt
```

```{python setPythonOptions}
pd.options.display.max_columns = None
pd.options.display.max_rows = None
```

```{python updateData}
# # Return all 3.7MM rows (as of Aug 4th, 2020)
# raw_data = pd.read_json('https://data.novascotia.ca/resource/3kpf-veux.json?$limit=4000000')
# 
# # Write file
# raw_data.to_csv('../00_data/ns_employee_absenteeism.csv')
```

```{python}
# desktop path
raw_data = pd.read_csv("D:/Personal/BIDAMIA_GoogleDrive/Business Intelligence/Python/Data/ns_employee_absenteeism.csv")
#raw_data = pd.read_csv("C:/Personal/BIDAMIA_GoogleDrive/Business Intelligence/Python/Data/ns_employee_absenteeism.csv")
# raw_data = pd.read_csv('../00_data/ns_employee_absenteeism.csv')
```


# Summary

Way back in August 2017 I started to apply R to working with data and one of my early projects was to explore the Nova Scotia (NS) Government employee absentee [data.](https://data.novascotia.ca/Public-Service/Nova-Scotia-Government-Employee-Absenteeism/3kpf-veux) You can see the original post [here.](https://bidamia.ca/post/nova-scotia-government-employee-absentee-analysis/)

I am revisiting this analysis because I am now learning python and I got tired of doing courses and wanted to apply what I learned so far. Also, there is now a few years more worth of data. 

Goal:

The code for this analysis can be found on my [Github page.](https://github.com/chrisselig/ns_employee_absenteeism_python)


# The Data

As before, I will start with an exploratory analysis of the data, but first, let's review the columns in the dataset.

1. absence_date: The date of absence for the employee

2. absence_hours: The number of hours the employee was absent

3. absence_type: The reason for the absence

4. absence_type_category: The absence reasons are grouped into standard categories

5. age_cohort_on_absence_date: The age range of the employee, the day they were absent

6. employee_type: Whether the employee was union, or non-union

7. gender: The gender of the employee

In total, there are 7 columns and 3,568,910 rows with data spanning from April 1st 2014 to March 31st, 2020.

```{python, results="asis"}
head_table = raw_data[['absence_date','absence_type_category','absence_type','absence_hours', 'employee_type']].head()[1:6]

head_table.to_html(index = False, justify='center')
```

# Data Cleaning section
fix 65+
remove time
change to date column?


# Exploratory Analysis

Originally, I started by taking a look at the total number of days absent which not only did not account for the fact that the people were not always absent for full days but I also only looked at total absences. A more realistic approach would have been to standardize absences per 100 people or absence hours per 100 people. Unfortunately, I cannot find any information on how many people are employed by the NS so I will continue with counts alone. 

```{python}
# raw_data.groupby(['age_cohort_on_absence_date']).value_counts
pivot = (raw_data.pivot_table(index=['age_cohort_on_absence_date'],values = ['absence_date'],aggfunc='size', fill_value=0)
    .sort_values(ascending=False))

# plt.bar()
# test.sort_values((['age_cohort_on_absence_date'], ascending = False))
```

```{python}
fig, ax = plt.subplots()
ax.spines['top'].set_visible(False)
ax.spines['right'].set_visible(False)
ax.spines['bottom'].set_visible(False)
ax.spines['left'].set_visible(False)

barh_absences_by_age_group = raw_data['age_cohort_on_absence_date'].value_counts().sort_values(ascending = True).plot(kind='barh',figsize = (7,6))

plt.title('Count of Absences by Age Group Cohort', y = 1.02)
plt.ylabel('Age Cohort')

plt.show()
```

```{python}

```


Check for statistical differences with union vs non-union employees
